---
title: "02_AssignASVs.Rmd"
author: "Emily Wahly"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
  code_folding: show
theme: spacelab
highlight: tango
toc: yes
toc_float:
  collapsed: no
smooth_scroll: yes
toc_depth: 3
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center", 
                      # Send figures generated in this file to this folder below
                      fig.path = "/local/workdir/erw98/Brassica_Growth_2026/figures/02_AssigningASVs")
knitr::opts_knit$set(root.dir = '/local/workdir/erw98/Brassica_Growth_2026/')

```
# Goals 

1. Infer an error model for in the filtered sequences, separately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately by applying the error model.
3. Merge forward and reverse ASVs into "contiguous ASVs".
4. Generate the first draft of ASV count table.
5. Quality Trimming of ASV lengths.
6. Remove chimeras. 
7. Assign Taxonomy with Silva Database. 
8. Write out relevant files: `asv_table`, `asvs_fasta`, `tax_table`, and `sample_data`.
9. Create a **S4** R data object, which in this case is known as a phyloseq, object from the `asv_table`, `tax_table`, and metadata. This will give us one data object to work with moving forward. :)

## Input 

1. Filtered fastq files generated from `01_QualityTrimming.Rmd`.
2. Sample Name vector.

## Output 

1. ASV Count Table: `asv_table` (with and without sequence names)
2. ASV fasta file: `asvs_fasta` for building a phylogenetic tree at a later step.
3. Taxonomy Table  `tax_table`
4. Sample Information: `sample_data`  track the reads lots throughout DADA2 workflow. 

# Set up the Environment 

## Set Seed & Threads

We need to be mindful of others when we are using the server. We learned in this lesson that we can use `top` and `htop` in the shell to see how many users are currently on the server. So, if you're working through something outside of class, it's ok to increase the number of threads. And, it will make everything go faster! 

Then, when we have a `multithread` parameter, we can set it to be `multithread = n_threads`. If we notice more people on the server, we can modify the n_threads in this one location. 

```{r set-seed}
# Set the seed for reproducibility
set.seed(238428)

# Let's make a parameter to set the number of threads 
n_threads = 30
```

## Timing of this script

Let's record how long this file took to run on the class server, which we will record at the end of the script. 

```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```


## Load Packages 
```{r load-packages}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, devtools, dada2, phyloseq, 
               patchwork, DT, install = FALSE)
```
#Load filtered fastq files
```{r load-files}
# Place filtered seq files into a variable 
filtered_fastqs_path <- "/local/workdir/erw98/Brassica_Growth_2026/data/01_Quality_Trimming/filtered_fastqs"
# Create Forward vector 
filtered_forward_reads <- 
  list.files(filtered_fastqs_path, pattern = "R1_filtered.fastq.gz",
             full.names = TRUE)
# Reverse vector 
filtered_reverse_reads <- 
    list.files(filtered_fastqs_path, pattern = "R2_filtered.fastq.gz",
             full.names = TRUE)

# Check 
filtered_forward_reads[1:5]
filtered_reverse_reads[1:5]

# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forward_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)
```
sample names
#Learn the Errors
MiSeq Runs: 40 Phred Scores

```{r learn-plot-errors}
# Forward Reads 
error_forward_reads <- 
  learnErrors(filtered_forward_reads, multithread = n_threads)
#81 samples
# Forward Error Plot 
forward_error_plot <- 
  plotErrors(error_forward_reads, nominalQ = TRUE) + 
  labs(title = "Forward Reads: Error Model")

# Reverse Reads 
error_reverse_reads <- 
  learnErrors(filtered_reverse_reads, multithread = n_threads)

# Reverse Error Plot 
reverse_error_plot <- 
  plotErrors(error_reverse_reads, nominalQ = TRUE) + 
  labs(title = "Reverse Reads: Error Model")

# Look at the plots together 
forward_error_plot + reverse_error_plot
```
#Infer ASVs
```{r infer-asvs, include=FALSE}
# Infer ASVs on the forward sequences
dada_forward <- 
  dada(filtered_forward_reads, 
       err = error_forward_reads,
       multithread = n_threads) 
# Take a look at the data
# What type of data structure is it? 
typeof(dada_forward) # It's a list 
length(dada_forward) # How big is it? One per sample!
#250 samples (500/2)...earlier it said 81?
dada_forward$'SRR11917050_R1_filtered.fastq.gz'
# Reverse ASVs
dada_reverse <- 
  dada(filtered_reverse_reads,
       err = error_reverse_reads ,
       multithread = n_threads)
# Check data 
dada_reverse[30]

```
#Merge Forward and Reverse ASVs
```{r merge}
merged_ASVs <- 
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)
# Evaluate the data output 
typeof(merged_ASVs) # A list
length(merged_ASVs) # Length of the number of samples! 252
head(names(merged_ASVs)) # Here, we can access our current sample names
# Inspect further for each sample
#head(merged_ASVs, n = 2) # A dataframe for each sample
# We have a dataframe in each part of our list! What are in the columns? 
glimpse(merged_ASVs$`SRR11917050_R1_filtered.fastq.gz`)
```
#Create Raw ASV Count Table

```{r asv-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Intuition Check: Type and dimensions of the data
dim(raw_ASV_table)   #250 2946 / now it says 252 18734?
typeof(raw_ASV_table) #integer
class(raw_ASV_table) #matrix/array
```
#Assess ASV Quality
INTERPRETATION #1: What is the hypothesized ASV length for your project data set? Please insert a description here, similar to the statements above, of the expected ASV length. Please provide a clear explanation using the following, which an example is provided above: (1) sequencing strategy, (2) total length of starting amplicons, (3) the expected ASV length without the primers, (4) length expected after the filterAndTrim() step, and (5) the expected percent overlap.

/They sequenced 300bp despite only using the 515F/806R region. ???

#ASV Length Stats
```{r stats}
# Calculate summary stats
# Longest ASV?
maxLength_ASV <- max(nchar(getSequences(raw_ASV_table))) #recall number characters nchar command
maxLength_ASV #510 / now 413
# Shortest ASV?
minLength_ASV <- min(nchar(getSequences(raw_ASV_table)))
minLength_ASV #290
# Mean ASV length?
meanLength_ASV <- mean(nchar(getSequences(raw_ASV_table))) 
meanLength_ASV #307
# Median ASV length?
medianLength_ASV <- median(nchar(getSequences(raw_ASV_table))) 
medianLength_ASV #292
# Create a table to Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table)))
```
#ASV Length Plot
```{r length plot}
# Inspect the distribution of sequence lengths of all ASVs in data set 
# AFTER TRIM
plot_ASVLength_raw <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Raw ASV Length (bps)")

# Show the plot
plot_ASVLength_raw
```
#Trim ASV Lengths
```{r trimming}
# Subset only ASVs that are 292 bps long 
raw_ASV_table_trimmed <- 
  raw_ASV_table[,nchar(colnames(raw_ASV_table)) == 292]

# Inspect the distribution of sequence lengths of all ASVs in dataset 
table(nchar(getSequences(raw_ASV_table_trimmed)))
# What proportion of total ASV sequences are left in the data? 
percRetained_Trimmed <- sum((raw_ASV_table_trimmed)/sum(raw_ASV_table)*100)
percRetained_Trimmed # 79.859% are retained... not as good as the 99% in the example
#83.62% Now!
# Inspect the distribution of sequence lengths of all ASVs in dataset 
# AFTER TRIM
plot_ASVLength_trimmed <- 
  data.frame(Seq_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = Seq_Length )) + 
  geom_histogram() + 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs", x = "Trimmed ASV Length (bps)")

# Show the plot 
plot_ASVLength_trimmed
#uhhh my peak is not at 3000 ASVs... visible in dim earlier... uhhh ok?
```
#Remove Chimeras
```{r chimeras}
# Remove the chimeras in the raw ASV table
noChimeras_ASV_table <- 
  removeBimeraDenovo(raw_ASV_table_trimmed, 
                     method="consensus", 
                     multithread = n_threads, 
                     verbose=TRUE)
#Identified 590 bimeras out of 1984 input sequences.
#this seems high??
dim(noChimeras_ASV_table) # 250 1394

# What proportion is left of the sequences? 
# Chimera removal compared to trimming 
percRetained_chimerasTrimmed <- sum(noChimeras_ASV_table)/sum(raw_ASV_table_trimmed)
# Chimera removal compared to raw  
percRetained_chimerasRaw <-sum(noChimeras_ASV_table)/sum(raw_ASV_table)
percRetained_chimerasRaw
percRetained_chimerasTrimmed
# Plot it 
plot_ASVLength_NoChimeras <- 
  data.frame(Seq_Length_NoChim = nchar(getSequences(noChimeras_ASV_table))) %>%
  ggplot(aes(x = Seq_Length_NoChim )) + 
  geom_histogram()+ 
  # include the x-axis scales
  scale_x_continuous(limits = c(0, maxLength_ASV + 5)) + 
  labs(y = "Number of ASVs \n (Post-Chimera Removal)", 
       x = "ASV Length (bps)")

# Show the plot
plot_ASVLength_NoChimeras #ouch
```

#Plot ASV Lengths after Chimera Removal
```{r plot-lengths}
plot_ASVLength_raw + plot_ASVLength_trimmed + plot_ASVLength_NoChimeras + 
    plot_annotation(tag_levels = 'A')
#so fun! fave plot
```

#Track Read Counts
```{r}
# A little function to identify number seqs 
getN <- function(x) sum(getUniques(x))

# Make the table to track the seqs 
track <- cbind(sapply(dada_forward, getN),
               sapply(dada_reverse, getN),
               sapply(merged_ASVs, getN),
               rowSums(noChimeras_ASV_table))

head(track)
#ok
# Update column names to be more informative (most are missing at the moment!)
colnames(track) <- c("denoisedF", "denoisedR", "merged", "nochim")
rownames(track) <- row.names(noChimeras_ASV_table)

# Generate a dataframe to track the reads through our DADA2 pipeline
track_counts_df <- 
  track %>%
  # make it a dataframe
  as.data.frame() %>%
  rownames_to_column(var = "sample_names")

# Now let's add a column for the number of ASVs
# First, intuition check that the samples match 
stopifnot(track_counts_df$sample_names == row.names(noChimeras_ASV_table))

# Now, let's add a new column with the number of ASVs
track_counts_df <- 
  track_counts_df %>%
  mutate(num_ASVs = rowSums(noChimeras_ASV_table > 1))

# Visualize it in table format 
DT::datatable(track_counts_df)

#cool... plot it?
track_counts_df %>%
  pivot_longer(denoisedF:nochim, names_to = "read_type", values_to = "num_reads") %>%
  mutate(read_type = fct_relevel(read_type, "denoisedF", "denoisedR", "merged", "nochim")) %>%
  ggplot(aes(x = read_type, y = num_reads, fill = read_type)) + 
  geom_line(aes(group = sample_names), color = "grey") + 
  geom_point(shape = 21, size = 3, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") + 
  labs(x = "Filtering Step", y = "Number of Sequences") + 
  theme_bw()
#well... uhh

#plot it
plot_ReadDepth <- 
  track_counts_df %>%
  ggplot(aes(x = nochim)) + 
  geom_histogram() + 
  labs(x = "Total # of Sequences", y = "# of Samples") + 
  theme_bw()

# What is the ASV richness per sample? 
plot_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = num_ASVs)) + 
  geom_histogram() + 
  labs(x = "Total # of ASVs", y = "# of Samples") + 
  theme_bw()

# Now, let's look at the relationship of ASVs and Sequencing depth 
plot_ReadDepth_ASVRichness <- 
  track_counts_df %>%
  ggplot(aes(x = nochim, y = num_ASVs)) + 
  geom_point() + 
  labs(x = "Total # of Sequences", y = "# of ASVs") + 
  theme_bw()

# Show the plots together 
plot_ReadDepth + plot_ASVRichness + plot_ReadDepth_ASVRichness + 
    plot_annotation(tag_levels = 'A')
#well that just doesn't look good.
```
#Assign Taxonomy
Here, we will use the silva database version 138.2, which has been properly formatted for DADA2. The files came from this DADA2-formatted reference databases website, which hosts several other reference taxomoy files for several popular taxonomic databases.
```{r taxonomy}
# Assign up to genus level 
taxa_train <- 
  assignTaxonomy(noChimeras_ASV_table, 
                 refFasta = "/workdir/in_class_data/taxonomy/silva_nr99_v138.2_toGenus_trainset.fa.gz", 
                 multithread = n_threads)

# Add the genus/species information 
taxa_addSpecies <- 
  addSpecies(taxa_train, 
              refFasta = "/workdir/in_class_data/taxonomy/silva_v138.2_assignSpecies.fa.gz")

# Inspect the taxonomy 
glimpse(taxa_addSpecies) # Note that the rownames are the ASV sequences!\# Let's removing the ASV sequence rownames for display only
taxa_print <- taxa_addSpecies 
rownames(taxa_print) <- NULL
head(taxa_print)
```
#EXPORTING
##1. ASV Tables
Let’s check that the ASVs are in the rows and the sample names are in the columns. If not, we can use the transpose function t() in R.
```{r}
# What's the current format of the ASV table?
head(rownames(noChimeras_ASV_table)) # Samples!
head(colnames(noChimeras_ASV_table)) # ASV Sequences
# Therefore, we need to transpose the matrix 
final_ASV_table_withSeqs <- t(noChimeras_ASV_table)

# Intuition check
head(rownames(final_ASV_table_withSeqs)) # ASV Sequences
head(colnames(final_ASV_table_withSeqs)) # Sample names
# Now, add a break in the script break if this isn't true! 
# Let's make sure the sample names match the file names in the matrix.
stopifnot(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1) == sample_names)

# Now, we've done some checks to prove to ourselves there will be no silent errors, 
# Let's rename! 
colnames(final_ASV_table_withSeqs) <- sample_names
head(colnames(final_ASV_table_withSeqs))
```


#### Rename ASVs

Then, we can also fix the names of our ASVs for our second ASV table where we will replace the ASV sequence names with ASV_1, ASV_2

```{r prepare-ASVcount-Seqtable}
# Give headers more manageable names
# First pull the ASV sequences from the rownames
ASV_seqs <- rownames(final_ASV_table_withSeqs)
ASV_seqs[1:5]

# How many ASVs? 
num_ASVs <- dim(final_ASV_table_withSeqs)[1] # select the number of rows
num_ASVs 

# Make an empty vector the length of the number of ASVs, 
# which is where we will place the new operational ASV names 
ASV_headers <- vector(num_ASVs, mode = "character")

# Let's mae sure we have an empty vector!
ASV_headers[1:5]
length(ASV_headers) # looks good! 

# Now, let's create a vector with ASV numbers
# loop through vector and fill it in with ASV names 
for (i in 1:num_ASVs) {
  # Add leading zero to ASV name so they print in correct order.
  ASV_number <- sprintf("%04d", i)
  # Now, rename each spot in the ASV header vector as we loop through the for loop
  ASV_headers[i] <- paste(">ASV", ASV_number, sep = "_")
}

save(final_ASV_table_withSeqs, file = "./raw_physeq")
raw_physeq <- final_ASV_table_withSeqs
```


### Names in ASV tables

#### Fix Sample Names 

Notice above that the sample names are actually file names. Let's fix this! 

> WARNING! This is an easy place where a silent error can step in where we could misname our samples, leading to catastrophe down the road. So, we need to be very careful and complete our checks and balances so we ensure the samples are not mixed up in the renaming process.

Reminder of output goals:
      1. ASV Count Table: asv_table (with and without sequence names)
      2. ASV fasta file: asvs_fasta for building a phylogenetic tree at a later step.
      3. Taxonomy Table tax_table
      4. Sample Information: sample_data track the reads lots throughout DADA2 workflow.

```{r SampleNames-ASV-table}
# Remember at the top of the file we created a vector of sample names 
head(sample_names)
# Let's check with the actual column names 
head(colnames(final_ASV_table_withSeqs)) # Sample names
# And then apply our sample name script to check, too
head(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1)) # Looks good! 

# Now, add a break in the script break if this isn't true! 
# Let's make sure the sample names match the file names in the matrix.
stopifnot(sapply(strsplit(colnames(final_ASV_table_withSeqs), "_"), `[`,1) == sample_names)

# Now, we've done some checks to prove to ourselves there will be no silent errors, 
# Let's rename! 
colnames(final_ASV_table_withSeqs) <- sample_names
head(colnames(final_ASV_table_withSeqs))
```


#### Rename ASVs

Then, we can also fix the names of our ASVs for our second ASV table where we will replace the ASV sequence names with ASV_1, ASV_2

```{r prepare-ASVcount-Seqtable}
# Give headers more manageable names
# First pull the ASV sequences from the rownames
ASV_seqs <- rownames(final_ASV_table_withSeqs)
ASV_seqs[1:5]

# How many ASVs? 
num_ASVs <- dim(final_ASV_table_withSeqs)[1] # select the number of rows
num_ASVs 

# Make an empty vector the length of the number of ASVs, 
# which is where we will place the new operational ASV names 
ASV_headers <- vector(num_ASVs, mode = "character")

# Let's mae sure we have an empty vector!
ASV_headers[1:5]
length(ASV_headers) # looks good! 

# Now, let's create a vector with ASV numbers
# loop through vector and fill it in with ASV names 
for (i in 1:num_ASVs) {
  # Add leading zero to ASV name so they print in correct order.
  ASV_number <- sprintf("%04d", i)
  # Now, rename each spot in the ASV header vector as we loop through the for loop
  ASV_headers[i] <- paste(">ASV", ASV_number, sep = "_")
}
ASV_headers[1:5]


# Create a new ASV table, which will have the ASV numbers as names 
# View(noChimeras_ASV_table) # To view the table
final_ASV_table <- final_ASV_table_withSeqs
glimpse(final_ASV_table)

## Replace the ASV seqs with the ASV numbers 
row.names(final_ASV_table) <- sub(">", "", ASV_headers)
final_ASV_table[1:5, 1:5]
#View(final_ASV_table) # To view the table
``` 

### Write the ASV Tables!

1. **With ASV seqs:** ASV headers include the *entire* ASV sequence 245 bases.
2. **with ASV names:** This includes re-written and shortened headers like ASV_1, ASV_2, etc, which will match the names in our fasta file below.

```{r write-asv-tables}
# 1. Write count table with ASV sequence names
write.table(final_ASV_table_withSeqs, 
            file = "/local/workdir/erw98/Brassica_Growth_2026/data/02_AssigningASVs/ASV_table_withSeqNames.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)

# 2. Write count table with ASV numbered names (e.g. ASV_1, ASV_2, etc)
write.table(final_ASV_table, 
            file = "/local/workdir/erw98/Brassica_Growth_2026/data/02_AssigningASVs/ASV_table.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```


## 2. ASV Fasta File 

Now, let's create a fasta file that has our ASV numbers with each of their corresponding ASV sequences. This file needs to be created because we will use it later to build a phylogenetic tree. 

**Why do we need a phylogenetic tree?** Many widely used microbial ecology metrics rely on phylogenetic relationships to quantify community similarity. For example, there are phylogenetic alpha-diversity metrics like the [phylogenetic Hill Numbers](https://www.annualreviews.org/content/journals/10.1146/annurev-ecolsys-120213-091540) and also beta-diversity measures that incorporate the branch lenghts of phylogenetic trees as it relates to community dissimimlarity, [like the abundance-unweighted and abundance-weighted UniFrac dissimilarity metrics](https://journals.asm.org/doi/full/10.1128/aem.71.12.8228-8235.2005). These metrics consider species presence and their evolutionary distances, conveying relationships of communities through an ecological *and* evolutionary framework. 

### Write the ASV Fasta File

```{r write-asv-fastas}
# Let's take our asv_headers
head(ASV_headers, n = 2)
# And combine it with their sequences
head(ASV_seqs, n = 2)

# Combine in a fasta format with the cbind() function
ASV_fasta <- c(rbind(ASV_headers, ASV_seqs))
head(ASV_fasta, n = 4)

# Then, let's write it to a fasta file!
# This will be our reference later on for which seq matches which ASV
write(ASV_fasta, "/local/workdir/erw98/Brassica_Growth_2026/data/02_AssigningASVs/ASVs.fasta")
```

## 3. Taxonomy Table 

Before we actually write the taxonomy table, let's put the ASV sequences as another column in our taxonomy table and then replace the rownames to be our numbered ASV names. 

### Reformat Taxonomy 

1. Move ASV sequences to column 
2. Replace ASV rownames with ASV numbered names

> WARNING! This is another easy place where a silent error can step in!! Here, we could accidentally misname our ASVs. So, we need to be very careful and perform intuition checks to ensure that the ASV names are not mixed up in the renaming process. 

```{r reformat-tax-table}
# Inspect the taxonomy table
dim(taxa_addSpecies) # ASVs are in rows and Kingdom, Phylum, etc in Columns 
colnames(taxa_addSpecies) # Column names are Linnean Taxonomy 
head(rownames(taxa_addSpecies), n = 2) # ASV names are rownames 
class(taxa_addSpecies) # Character matrix

##### Prepare tax table 
# 1. Add the ASV sequences from the rownames to a column 
new_tax_table <- 
  taxa_addSpecies%>%
  as.data.frame() %>%
  rownames_to_column(var = "ASVseqs") 

# Intuition check 
glimpse(new_tax_table)

# IMPORTANT! Let's do our intuition check 
# This is where we ensure we don't mix up the ASV names!
stopifnot(new_tax_table$ASVseqs == rownames(final_ASV_table_withSeqs))

# Now let's add the ASV names 
rownames(new_tax_table) <- rownames(final_ASV_table)
head(new_tax_table)

### Final prep of tax table. Add new column with ASV names 
ASV_tax_table <- 
  new_tax_table %>%
  # add rownames from count table for phyloseq handoff
  mutate(ASV = rownames(final_ASV_table)) %>%
  # Reorder the columns
  dplyr::select(Kingdom, Phylum, Class, Order, Family, Genus, Species, ASV, ASVseqs)

# Assign the rownames, which is required by phyloseq
rownames(ASV_tax_table) <- ASV_tax_table$ASV

# Take a quick look
glimpse(ASV_tax_table)

# Intution check
stopifnot(ASV_tax_table$ASV == rownames(ASV_tax_table), 
          rownames(ASV_tax_table) == rownames(ASV_tax_table))
```

### Write the Taxonomy Table 

Now, let's write out the taxonomy table 

```{r write-tax-table}
# Write the table 
write.table(ASV_tax_table, 
            file = "/local/workdir/erw98/Brassica_Growth_2026/data/02_AssigningASVs/ASV_taxonomy.tsv", 
            sep = "\t", quote = FALSE, col.names = NA)
```


## 4. DADA2 Sample Data 

Let's save the `track_counts_df`, which can be useful for downstream analyses. 

```{r save-track reads}
# And save the track_counts_df a R object, which we will merge with metadata information in the next step of the analysis in analysis/02_Taxonomic_Assignment. 
save(track_counts_df, file = "/local/workdir/erw98/Brassica_Growth_2026/data/02_AssigningASVs/track_read_counts.RData")
```




# Phyloseq S4 Object Creation 

To start this section of this lesson, please make sure that you have phyloseq installed! 

## 5. Metadata
```{r make-metadata}
# 1. Import the CSV file
Sra_df <- read.csv("/local/workdir/erw98/Brassica_Growth_2026/SraRunTable.csv")
View(final_ASV_table)
rownames(Sra_df) <- Sra_df$Run
metadata_physeq <- sample_data(Sra_df)
sample_names(metadata_physeq)
colnames(final_ASV_table)

stopifnot(all(sample_names(metadata_physeq) == colnames(final_ASV_table)))
#troubleshooting this stop:
length(sample_names(metadata_physeq))
length(colnames(final_ASV_table))
#see a difference in # of samples
head(sample_names(metadata_physeq))
head(colnames(final_ASV_table))
#find the overlap
intersect_samples <- intersect(
  sample_names(metadata_physeq),
  colnames(final_ASV_table)
)

length(intersect_samples)
#252 is the number I want. Which ones are overlap?
setdiff(colnames(final_ASV_table), sample_names(metadata_physeq))  # in ASV only
setdiff(sample_names(metadata_physeq), colnames(final_ASV_table))  # in metadata only

#intersect them
# Subset ASV table
final_ASV_table <- final_ASV_table[, intersect_samples]

# Subset metadata
metadata_physeq <- metadata_physeq[intersect_samples, ]

#check again 
stopifnot(all(sample_names(metadata_physeq) == colnames(final_ASV_table)))
length(filtered_forward_reads)
length(filtered_reverse_reads)
all(basename(filtered_forward_reads) %in% basename(filtered_reverse_reads))

get_sample_id <- function(x) {
  sub("_R[12]_filtered.fastq.gz", "", basename(x))
}

forward_ids <- get_sample_id(filtered_forward_reads)
reverse_ids <- get_sample_id(filtered_reverse_reads)
all(forward_ids %in% reverse_ids)
all(reverse_ids %in% forward_ids)
# Reorder reverse reads to match forward reads
filtered_reverse_reads <- filtered_reverse_reads[
  match(forward_ids, reverse_ids)
]

# Final safety check
stopifnot(all(get_sample_id(filtered_forward_reads) ==
              get_sample_id(filtered_reverse_reads)))
head(forward_ids)
head(reverse_ids)

#rebuild phyloseq object
raw_physeq <- phyloseq(
  otu_table(final_ASV_table, taxa_are_rows = TRUE),
  tax_table(as.matrix(ASV_tax_table)), 
  metadata_physeq
)
# 2. Save the object to an .RData file
save(Sra_df, file = "/local/workdir/erw98/Brassica_Growth_2026/metadata.RData")

# read the SRA run table into R. This is my equivalent of the metadata, right?
load("/local/workdir/erw98/Brassica_Growth_2026/metadata.RData")
Sra_df
# The command above will load in a new object called:
#rename it:
metadata_physeq <- Sra_df
View(metadata_physeq)
# Check it out
sample_names(metadata_physeq)
# how many samples?
length(sample_names(metadata_physeq))
```


```{r make-phyloseq}
# Construct the phyloseq object 
raw_physeq <- phyloseq(otu_table(final_ASV_table, taxa_are_rows = TRUE),
                       # The taxonomy table needs to first be converted into a matrix
                       tax_table(as.matrix(ASV_tax_table)), 
                       # Next, add the sample_data
                       metadata_physeq)

# Check the phyloseq object!
raw_physeq

# Now, save the phyloseq object, but use the correct path to your folder:
save(raw_physeq, file = "/local/workdir/erw98/Brassica_Growth_2026/data/raw_physeq.RData")
```

Now that we have saved our data files, we are good to go! 

# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```